---
title: "data_exploration"
output: pdf_document
date: "2025-03-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

```{r}
library(dplyr)        # A package for data manipulation
library(sf)           # Simple feature for R
library(spdep)        # Functions and tests for evaluating spatial patterns 
library(tidyr)        # Tools to create tidy data
library(INLA)         # Integrated Nested Laplace Approximation package
library(ggplot2)      # A package for creating maps and graphs
library(viridis)      # A package providing color palettes 
library(patchwork)    # A package to compose plots

# For tables in RMarkdown
library(knitr)
library(kableExtra)
```

```{r}
# rstudioapi::getSourceEditorContext()$path

RESP_DATA = read.csv("/Users/jacksonjoseph/Desktop/HDA/Term2/Bayesian/mini_proj/dat/Data_final.csv")
shape_file = st_read("/Users/jacksonjoseph/Desktop/HDA/Term2/Bayesian/mini_proj/dat/LocalAuthorities.shp")
```
```{r}
RESP_DATA
```
```{r}
# we consider yearly pneumonia deaths the equivalent of 'hospital admissions for respiratory conditions in greater glasgow and clyde'

# we start by computing the total number of cases of deaths per year, and format output in a table. 

kable(RESP_DATA %>%
        group_by(Year_death) %>%
         summarise(Y = sum(Y), expected=sum(E), booktabs = T, caption = "Hospital admissions by year")) %>% 
    kable_styling(bootstrap_options = "striped", full_width = F, position = "center")

```

```{r}
# now plot the spatial object using ggplot2: 

ggplot() + 
      geom_sf(data = shape_file, color = "blue", fill = "white") + 
      coord_sf() +    #axis limits and CRS
      theme_bw() +    # dark-on-light theme
      theme(axis.title = element_text(size = 14),
            axis.text = element_text(size = 12))
```


```{r}
# in order to use the same data structure for both the space-only model and later the space-time model, new set of data is formed 
# by aggregating both observed and expected counts over time, poisson-log linear model then fitted, assuming BYM2 model for random effects 
# define neighbors and create weights list 

shape_nb = poly2nb(shape_file, snap = 1000, queen = TRUE)
summary(shape_nb)
```
```{r}
# now convert list of neighbors to inla format using nb2WB

nb2INLA("shape.graph", shape_nb)
shape.adj = paste(getwd(), "/shape.graph", sep = "")
```

```{r}
# aggregate observed and expected cases over geographical areas 

RESP_DATAagg = RESP_DATA %>% group_by(lad09cd) %>%
  summarize(observed = sum(Y), 
            expected = sum(E)) %>% 
  dplyr::rename(O = observed, E = expected)

RESP_DATAagg = RESP_DATAagg %>% mutate(SMR = O/E)
```

```{r}
RESP_DATAagg
```


```{r}
# produce a spatial map of aggregated SMRs using ggplot2. 

RESP_DATAagg$SMRcat = cut(RESP_DATAagg$SMR, 
                      breaks=c(min(RESP_DATAagg$SMR), 
                               0.4, 0.6, 0.8, 1, 1.2, 1.4, 1.6, 
                               max(RESP_DATAagg$SMR)), include.lowest = T)

map_SMR = left_join(shape_file, RESP_DATAagg, by = c("lad09cd" = "lad09cd"))
```

```{r}
ggplot() + geom_sf(data = map_SMR, col = NA) + aes(fill = SMRcat) +
  theme_bw() + scale_fill_viridis_d() + 
  guides(fill=guide_legend(title="SMR")) 

```

```{r}
map_SMR
```

```{r}
nrow(RESP_DATAagg)
```



```{r}
# fit the hierarchical poisson log-linear model in INLA. 
ID = seq(1,322)
formula_BYM2 = O ~ f(ID, model="bym2", graph=shape.adj,
                            hyper=list(prec = list(
        prior = "pc.prec",
        param = c(0.5 / 0.31, 0.01)),
        phi = list(
        prior = "pc",
        param = c(0.5, 2 / 3))))    

sBYM.model = inla(formula=formula_BYM2, family="poisson", data=RESP_DATAagg, E=E, control.compute=list(waic=TRUE))
```

```{r}
# obtain posterior summary statistics mean/posterior probability that the residual is above 1) of the parameters of interest 

#Relative risks
RR_sBYM = c()

for(i in 1:322){
  RR_sBYM[i] = inla.emarginal(function(x) exp(x), 
        sBYM.model$marginals.random$ID[[i]])
}

#Posterior probabilities
RR_sBYM_marg = sBYM.model$marginals.random$ID[1:322]
PP_sBYM = lapply(RR_sBYM_marg, function(x) {1-inla.pmarginal(0,x)})
```

```{r}
# obtain the posterior estimates from the spatial model to be plotted, that is (i) area level posterior mean of residual RR 
# ii) the posterior probability that the residual RRs > 1

resRR_PP = data.frame(resRR=RR_sBYM, 
                       PP=unlist(PP_sBYM),
                      SP_ID=RESP_DATAagg[,1])
```

```{r}
# produce map of posterior mean of residual RRs and posterior probabilities that residual RRS are > 1

resRR_PP$resRRcat = cut(resRR_PP$resRR, breaks=c(min(resRR_PP$resRR), 
                  0.4, 0.6, 0.8, 1, 1.2, 1.4, 1.6, 
                  max(resRR_PP$resRR)),include.lowest = T)


# breakpoints
resRR_PP$PPcat = cut(resRR_PP$PP, c(0, 0.2, 0.8, 1.00), include.lowest = TRUE)

# join sf object and dataframe with posterior estimates. 

map_RR_PP = left_join(shape_file, resRR_PP, by = c("lad09cd" = "lad09cd"))


# produce maps 

p1 = ggplot() + geom_sf(data = map_RR_PP) + aes(fill = resRRcat) +
  theme_bw() + scale_fill_brewer(palette = "PuOr") + 
  guides(fill=guide_legend(title="RR")) + ggtitle("RR Spatial model") + 
  theme(text = element_text(size=15), 
                  axis.text.x = element_blank(), 
                  axis.text.y = element_blank(), plot.title = element_text(size = 12, face = "bold"))

p2 = ggplot() + geom_sf(data = map_RR_PP) + aes(fill = PPcat) +
  theme_bw() +
  scale_fill_viridis(
    option = "plasma", name="PP",
    discrete = T,
    direction = -1,
    guide = guide_legend(
      title.position = 'top',
      reverse = T
    )) +  ggtitle("PP Spatial model") + theme(text = element_text(size=15), 
                  axis.text.x = element_blank(), 
                  axis.text.y = element_blank(), plot.title = element_text(size = 12, face = "bold")) 

p1|p2

# the BYM2 model smooths raw SMRs by borrowing strength from neighbouring areas while still allowing for independent noise. The PC guards against overfitting

# first map tells you magnitude of excess or reduced risk after smoothing away random noise 

# second map is about how certain we are that RR > 1 i.e. if PP close to 0, then strong evidence that areas risk is actually below average, PP>0.8 strong evidence areas risk above average 

# therefore light and dark areas are strongest evidence for real excess risk (positive or negative)

```

```{r}
# estimate the spatial fraction 
# as bymw has structured and unstructured components useful to get ideas about strength of spatially structured componets to indicate clustering in data 

sBYM.model$summary.hyperpar # says about 0.4 of spatial variablity explained by spatiall structured component 
```

```{r}
RESP_DATA
```

now we do a spatio-temporal model - no interactions 
```{r}

# first prepare data, we join shapefile to make sure order is the same and create ID for time and one for space 

#Join the data with the shapefile so the order of the shapefile is maintained.  
RESP_DATA_ST = left_join(shape_file, RESP_DATA, by="lad09cd")

#Rename the columns of Observed and Expected as we did before
RESP_DATA_ST = RESP_DATA_ST  %>% dplyr::rename(O = Y, E = E)

#Create the ID for year (time)
RESP_DATA_ST$ID.time = RESP_DATA_ST$Year_death - 2002

#Create the ID for space
RESP_DATA_ST$ID.space = rep(seq(1,322),each=16)
```

```{r}

# run the model in INLA

formula_ST_noint = O ~ f(ID.space, model="bym2", graph=shape.adj,
                            hyper=list(prec = list(
                            prior = "pc.prec",
                            param = c(0.5 / 0.31, 0.01)),
                            phi = list(
                            prior = "pc",
                            param = c(0.5, 2 / 3)))) + f(ID.time,model="rw1", hyper=list(prec = list(
                            prior = "pc.prec",
                            param = c(0.5 / 0.31, 0.01))))
                            
stBYM.model = inla(formula=formula_ST_noint, family="poisson", 
                   data=RESP_DATA_ST, E=E, 
                   control.compute=list(waic=TRUE))


```

```{r}
# create the posterior mean for the spatial and temporal effects 

RR_stBYM = c()

for(i in 1:322){
  RR_stBYM[i] = inla.emarginal(function(x) exp(x), 
        stBYM.model$marginals.random$ID.space[[i]])
}

#Posterior probabilities (for spatial RR)
RR_stBYM_marg = stBYM.model$marginals.random$ID.space[1:322]
PP_stBYM = lapply(RR_stBYM_marg, function(x) {1-inla.pmarginal(0,x)})

#Temporal Relative risks and CI95
RR_stRW_RR = c()
RR_stRW_lo = c()
RR_stRW_hi = c()

for(i in 1:16){
  #Posterior mean
  RR_stRW_RR[i] = inla.emarginal(function(x) exp(x), 
        stBYM.model$marginals.random$ID.time[[i]])
  #2.5% quantile 
  RR_stRW_lo[i] = inla.qmarginal(0.025,inla.tmarginal(function(x) exp(x), stBYM.model$marginals.random$ID.time[[i]]))
  #97.5% quantile 
  RR_stRW_hi[i] = inla.qmarginal(0.975, inla.tmarginal(function(x) exp(x), stBYM.model$marginals.random$ID.time[[i]]))
}

RR_stRW = data.frame(RR=RR_stRW_RR,low=RR_stRW_lo,high=RR_stRW_hi)


```

```{r}
# plot the temporal residuals 
Temp1 = ggplot(RR_stRW, aes(seq(2002,2017), RR)) + geom_line() + ggtitle("ST model No Int") + geom_ribbon(aes(ymin=low,ymax=high), alpha=0.2) + labs(x="year")

Temp1
```
```{r}
typeof(shape.adj)
```


```{r}
# map the spatial residuals RRs with ggplot2. Compare this map of residual RR obtained from spatial model 
resRR_PP_st = data.frame(resRR=RR_stBYM, 
                       PP=unlist(PP_stBYM),
                      SP_ID=RESP_DATAagg[,1])
# breakpoints
resRR_PP_st$resRRcat = cut(resRR_PP_st$resRR, breaks=c(min(resRR_PP_st$resRR), 
                  0.4, 0.6, 0.8, 1, 1.2, 1.4, 1.6, 
                  max(resRR_PP_st$resRR)),include.lowest = T)

resRR_PP_st$PPcat = cut(resRR_PP_st$PP, c(0, 0.2, 0.8, 1.00), include.lowest = TRUE)

map_RR_ST = left_join(shape_file, resRR_PP_st, by = c("lad09cd" = "lad09cd"))


p3 = ggplot() + geom_sf(data = map_RR_ST) + aes(fill = resRRcat) +
  theme_bw() + scale_fill_brewer(palette = "PuOr") + 
  guides(fill=guide_legend(title="RR")) +  ggtitle("RR ST model") +
  theme(text = element_text(size=15), 
        axis.text.x = element_blank(), 
        axis.text.y = element_blank(), plot.title = element_text(size = 12, face = "bold")) 

p4 = ggplot() + geom_sf(data = map_RR_ST) + aes(fill = PPcat) +
  theme_bw() +
  scale_fill_viridis(
    option = "plasma",
    name = "PP ST model",
    discrete = T,
    direction = -1,
    guide = guide_legend(
      title.position = 'top',
      reverse = T
    )) +  ggtitle("PP ST model") + theme(text = element_text(size=15), 
        axis.text.x = element_blank(), 
        axis.text.y = element_blank(), plot.title = element_text(size = 12, face = "bold"))

(p1|p2) / (p3|p4)
```

```{r}
# next steps are to add deprivation and physical activity 

# make sure they are standardised 

RESP_DATA_ST = RESP_DATA_ST %>% 
  mutate(deprivation_s = scale(Deprivation), 
         physact_s  = scale(PhysicalActivity))

RESP_DATA_ST
```
```{r}
# update model formula 

formula_cov <- O ~ deprivation_s + physact_s +
  f(ID.space, model="bym2", graph=shape.adj,
    hyper=list(
      prec = list(prior="pc.prec", param=c(0.5/0.31, 0.01)),
      phi  = list(prior="pc",       param=c(0.5, 2/3))
    )) +
  f(ID.time, model="rw1",
    hyper=list(
      prec = list(prior="pc.prec", param=c(0.5/0.31, 0.01))
    ))
```

```{r}
# refit in INLA and compare WAIC
stBYM_cov <- inla(
  formula_cov,
  family="poisson",
  data=RESP_DATA_ST,
  E=E,
  control.compute=list(waic=TRUE)
)
stBYM_cov$waic$waic
```

```{r}
cat("No‑covariate WAIC =", stBYM.model$waic$waic, "\n",
    "Covariate‑adjusted WAIC =", stBYM_cov$waic$waic, "\n")
```


```{r}
fixed <- stBYM_cov$summary.fixed[, c("mean","0.025quant","0.975quant")]
fixed_rr <- exp(fixed)
rownames(fixed_rr) <- c("Intercept","Deprivation (1 SD)","Physical activity (1 SD)")
round(fixed_rr, 3)
```

```{r}
RR_cov <- sapply(stBYM_cov$marginals.random$ID.space, 
                 function(m) inla.emarginal(function(x) exp(x), m))
PP_cov <- sapply(stBYM_cov$marginals.random$ID.space, 
                 function(m) 1 - inla.pmarginal(0, m))
res_cov <- data.frame(lad09cd=unique(RESP_DATAagg$lad09cd), resRR=RR_cov, PP=PP_cov)

res_cov$resRRcat <- cut(res_cov$resRR, breaks=c(min(res_cov$resRR),0.4,0.6,0.8,1,1.2,1.4,1.6,max(res_cov$resRR)))
res_cov$PPcat <- cut(res_cov$PP, breaks=c(0,0.2,0.8,1), include.lowest=TRUE)

map_cov <- left_join(shape_file, res_cov, by="lad09cd")

p3 <- ggplot(map_cov) + geom_sf(aes(fill=resRRcat)) +
      theme_bw() + scale_fill_brewer(palette="PuOr") + 
      ggtitle("Residual RR — ST model w/ covariates")

p4 <- ggplot(map_cov) + geom_sf(aes(fill=PPcat)) +
      theme_bw() + scale_fill_viridis(discrete=TRUE, direction=-1) +
      ggtitle("Posterior P(RR>1) — ST model w/ covariates")

(p3|p4)

```

```{r}
# temporal analysis as well 

temporal <- data.frame(
  year = 2002:2017,
  RR   = sapply(stBYM_cov$marginals.random$ID.time, 
                function(m) inla.emarginal(function(x) exp(x), m)),
  low  = sapply(stBYM_cov$marginals.random$ID.time,
                function(m) exp(inla.qmarginal(0.025, m))),
  high = sapply(stBYM_cov$marginals.random$ID.time,
                function(m) exp(inla.qmarginal(0.975, m)))
)

ggplot(temporal, aes(year, RR)) +
  geom_line() +
  geom_ribbon(aes(ymin=low, ymax=high), alpha=0.2) +
  labs(title="Temporal residual RR — ST model w/ covariates", x="Year", y="Relative risk") +
  theme_minimal()
```
################################################################################
# Additional Spatio-Temporal Analysis - Type I Interaction 
################################################################################

Now, we extend the above spatio-temporal analysis to a separable space-time model with type I interaction: 

18. Specify the formula and run the model in `INLA`. Remember that you need to create an index which goes from 1 to the length of the dataset (i.e. the space x time). For the `iid` model defining the interaction term, use the PC prior previously used for the `rw1` model. Call the output `stIntI.BYM.model` (remember to monitor the WAIC)

```{r eval=TRUE}
RESP_DATA_ST$ID.space.time = seq(1,dim(RESP_DATA_ST)[1])

formula_ST_intI = O ~ f(ID.space, model="bym2", graph=shape.adj,
                            hyper=list(prec = list(
                            prior = "pc.prec",
                            param = c(0.5 / 0.31, 0.01)),
                            phi = list(
                            prior = "pc",
                            param = c(0.5, 2 / 3)))) + 
                      f(ID.time,model="rw1", hyper=list(prec = list(
                            prior = "pc.prec",
                            param = c(0.5 / 0.31, 0.01))))+
                      f(ID.space.time,model="iid", hyper=list(prec = list(
                            prior = "pc.prec",
                            param = c(0.5 / 0.31, 0.01))))
                    
                            
stIntI.BYM.model = inla(formula=formula_ST_intI, family="poisson", data=RESP_DATA_ST, E=E,
                        control.compute=list(dic=TRUE, waic=TRUE))

```

19. Create the posterior mean for the spatial and temporal effects and compare with the ST model results without interaction
```{r eval=TRUE} 
#Spatial Relative risks
RR_stIntI.BYM = c()

for(i in 1:322){
  RR_stIntI.BYM[i] = inla.emarginal(function(x) exp(x), 
        stIntI.BYM.model$marginals.random$ID.space[[i]])
}

#Posterior probabilities (for spatial RR)
RR_stIntI.BYM_marg = stIntI.BYM.model$marginals.random$ID.space[1:322]
PP_stIntI.BYM = lapply(RR_stIntI.BYM_marg, function(x) {1-inla.pmarginal(0,x)})

#Temporal Relative risks and CI95
RR_stIntI.RW_RR = c()
RR_stIntI.RW_lo = c()
RR_stIntI.RW_hi = c()

for(i in 1:16){
  #Posterior mean
  RR_stIntI.RW_RR[i] = inla.emarginal(function(x) exp(x), 
        stIntI.BYM.model$marginals.random$ID.time[[i]])
  #2.5% quantile 
  RR_stIntI.RW_lo[i] = inla.qmarginal(0.025,inla.tmarginal(function(x) exp(x), stIntI.BYM.model$marginals.random$ID.time[[i]]))
  #97.5% quantile 
  RR_stIntI.RW_hi[i] = inla.qmarginal(0.975, inla.tmarginal(function(x) exp(x), stIntI.BYM.model$marginals.random$ID.time[[i]]))
}

RR_stIntI.RW = data.frame(RR=RR_stIntI.RW_RR,low=RR_stIntI.RW_lo,high=RR_stIntI.RW_hi)
```	

20. Plot the temporal residual RRs (`RR_stWR`)
```{r eval=TRUE}

# NB: WHILE THIS IS A COMPARISON OF THE ST MOD W/ AND W/O TYPE I INTERACTION, THE Y-AXIS MUST BE FIXED FOR BOTH PLOTS TO FACILITATE ACTUAL COMPARISON

Temp2 = ggplot(RR_stIntI.RW, aes(seq(2002,2017), RR)) + geom_line() + ggtitle("ST model Int I") + geom_ribbon(aes(ymin=low,ymax=high), alpha=0.2) + labs(x="year")

Temp1 | Temp2

# Temp2 = ggplot(RR_stIntI.RW, aes(seq(2002,2017), RR)) + geom_line() + ggtitle("ST model Int I") + geom_ribbon(aes(ymin=low,ymax=high), alpha=0.2) + labs(x="year")
# 
# Temp2
```

21. Map the spatial residual RRs (`RR_stIntI.BYM`) with `ggplot2` package using the following breakpoints [min,0.4], (0.4-0.6], (0.6-0.8], (0.8,1], (1,1.2], (1.2-1.4], (1.4-1.6], (1.6-max]. 
Compare this map against the map of the residual RR obtained from the spatio-temporal model with no interaction. Try to comment the results

```{r eval=TRUE, echo=TRUE, include=TRUE}
resRR_PP_stIntI = data.frame(resRR=RR_stIntI.BYM, 
                       PP=unlist(PP_stIntI.BYM),
                      SP_ID=RESP_DATAagg[,1])
# breakpoints
resRR_PP_stIntI$resRRcat = cut(resRR_PP_stIntI$resRR, breaks=c(min(resRR_PP_stIntI$resRR), 
                  0.4, 0.6, 0.8, 1, 1.2, 1.4, 1.6, 
                  max(resRR_PP_stIntI$resRR)),include.lowest = T)

resRR_PP_stIntI$PPcat = cut(resRR_PP_stIntI$PP, c(0, 0.2, 0.8, 1.00), include.lowest = TRUE)

map_RR_ST.IntI = left_join(shape_file, resRR_PP_stIntI, by = c("lad09cd" = "lad09cd"))
```


```{r mapRRstIntI, eval=TRUE, echo=TRUE, include=TRUE, fig.cap="Spatio-temporal model: Map of the residual RRs and posterior probabilities", , fig.width=12, fig.height=10}
p5 = ggplot() + geom_sf(data = map_RR_ST.IntI) + aes(fill = resRRcat) +
  theme_bw() + scale_fill_brewer(palette = "PuOr") + 
  guides(fill=guide_legend(title="RR")) +  ggtitle("RR ST model Int I") +
  theme(text = element_text(size=15), 
        axis.text.x = element_blank(), 
        axis.text.y = element_blank(), plot.title = element_text(size = 12, face = "bold")) 

p6 = ggplot() + geom_sf(data = map_RR_ST.IntI) + aes(fill = PPcat) +
  theme_bw() +
  scale_fill_viridis(
    option = "plasma",
    name = "PP ST model Int I",
    discrete = T,
    direction = -1,
    guide = guide_legend(
      title.position = 'top',
      reverse = T
    )) +  ggtitle("PP ST model Int I") + theme(text = element_text(size=15), 
        axis.text.x = element_blank(), 
        axis.text.y = element_blank(), plot.title = element_text(size = 12, face = "bold"))

# (p1|p2) / (p3|p4) / (p5|p6)
```


22. Plot the space-time interaction

```{r include=TRUE, fig.width=20, fig.height=10}
RESP_DATA_ST$intI = stIntI.BYM.model$summary.random$ID.space.time$mean

RESP_DATA_ST$intI_cat = cut(RESP_DATA_ST$intI,  breaks=c(-1,-0.05, 
                  -0.01, 0.01, 0.05, 1),include.lowest = T)
ggplot() +
  geom_sf(data = RESP_DATA_ST, aes(fill = intI_cat))+ theme_bw() +  scale_fill_brewer(palette = "PuOr") + 
            guides(fill=guide_legend(title=NULL)) + 
            theme(text = element_text(size=20), 
                  axis.text.x = element_blank(), 
                  axis.text.y = element_blank()) +
facet_wrap(~ ID.time, ncol =8, labeller=labeller(ID.time=c("0"="2002","1"="2003","2"="2004","3"="2005","4"="2006","5"="2007","6"="2008","7"="2009","8"="2010","9"="2011", "10"="2012","11"="2013","12"="2014","13"="2015", "14"="2016","15"="2017"))) +
labs("")
```
We can see that there is not clear pattern in the interactions.

23. Get a table of the hyperparameters. How do you interpret this table?

```{r eval=TRUE, echo=TRUE}

dat.hyper2 = 
  round(
  data.frame(median = stIntI.BYM.model$summary.hyperpar[,4],
    LL = stIntI.BYM.model$summary.hyperpar[,3], 
    UL = stIntI.BYM.model$summary.hyperpar[,5]),
  digits = 3)

row.names(dat.hyper2) = 
  rownames(stIntI.BYM.model$summary.hyperpar)

knitr::kable(dat.hyper2, caption = "Posterior median and 95% CrI of hyperparameters.") %>%  kable_styling(bootstrap_options = "striped", full_width = F, position = "center")

```
################################################################################
SPATIAL TEMPORAL TYPE I WITH COVARS 
################################################################################

Now, we extend the above spatio-temporal analysis to a separable space-time model with type I interaction: 

18. Specify the formula and run the model in `INLA`. Remember that you need to create an index which goes from 1 to the length of the dataset (i.e. the space x time). For the `iid` model defining the interaction term, use the PC prior previously used for the `rw1` model. Call the output `stIntI.BYM.model` (remember to monitor the WAIC)

```{r eval=TRUE}
RESP_DATA_ST$ID.space.time = seq(1,dim(RESP_DATA_ST)[1])

formula_ST_intI_cov = O ~ deprivation_s + physact_s +
  f(ID.space, model="bym2", graph=shape.adj,
                            hyper=list(prec = list(
                            prior = "pc.prec",
                            param = c(0.5 / 0.31, 0.01)),
                            phi = list(
                            prior = "pc",
                            param = c(0.5, 2 / 3)))) + 
                      f(ID.time,model="rw1", hyper=list(prec = list(
                            prior = "pc.prec",
                            param = c(0.5 / 0.31, 0.01))))+
                      f(ID.space.time,model="iid", hyper=list(prec = list(
                            prior = "pc.prec",
                            param = c(0.5 / 0.31, 0.01))))
                    
                            
stIntI.BYM.model_cov = inla(formula=formula_ST_intI_cov, family="poisson", data=RESP_DATA_ST, E=E,
                        control.compute=list(dic=TRUE, waic=TRUE))

```

19. Create the posterior mean for the spatial and temporal effects and compare with the ST model results without interaction
```{r eval=TRUE} 
#Spatial Relative risks
RR_stIntI.BYM_cov = c()

for(i in 1:322){
  RR_stIntI.BYM_cov[i] = inla.emarginal(function(x) exp(x), 
        stIntI.BYM.model_cov$marginals.random$ID.space[[i]])
}

#Posterior probabilities (for spatial RR)
RR_stIntI.BYM_cov_marg = stIntI.BYM.model_cov$marginals.random$ID.space[1:322]
PP_stIntI.BYM_cov = lapply(RR_stIntI.BYM_cov_marg, function(x) {1-inla.pmarginal(0,x)})

#Temporal Relative risks and CI95
RR_stIntI_cov.RW_RR = c()
RR_stIntI_cov.RW_lo = c()
RR_stIntI_cov.RW_hi = c()

for(i in 1:16){
  #Posterior mean
  RR_stIntI_cov.RW_RR[i] = inla.emarginal(function(x) exp(x), 
        stIntI.BYM.model_cov$marginals.random$ID.time[[i]])
  #2.5% quantile 
  RR_stIntI_cov.RW_lo[i] = inla.qmarginal(0.025, inla.tmarginal(function(x) exp(x), stIntI.BYM.model_cov$marginals.random$ID.time[[i]]))
  #97.5% quantile 
  RR_stIntI_cov.RW_hi[i] = inla.qmarginal(0.975, inla.tmarginal(function(x) exp(x), stIntI.BYM.model_cov$marginals.random$ID.time[[i]]))
}

RR_stIntI_cov.RW = data.frame(RR=RR_stIntI_cov.RW_RR,low=RR_stIntI_cov.RW_lo,high=RR_stIntI_cov.RW_hi)
```	

20. Plot the temporal residual RRs (`RR_stWR`)
```{r eval=TRUE}

Temp3 = ggplot(RR_stIntI_cov.RW, aes(seq(2002,2017), RR)) + geom_line() + ggtitle("ST model Int I Covars") + geom_ribbon(aes(ymin=low,ymax=high), alpha=0.2) + labs(x="year")

Temp1 | Temp2 | Temp3

# Temp2 = ggplot(RR_stIntI.RW, aes(seq(2002,2017), RR)) + geom_line() + ggtitle("ST model Int I") + geom_ribbon(aes(ymin=low,ymax=high), alpha=0.2) + labs(x="year")
# 
# Temp2
```

21. Map the spatial residual RRs (`RR_stIntI.BYM`) with `ggplot2` package using the following breakpoints [min,0.4], (0.4-0.6], (0.6-0.8], (0.8,1], (1,1.2], (1.2-1.4], (1.4-1.6], (1.6-max]. 
Compare this map against the map of the residual RR obtained from the spatio-temporal model with no interaction. Try to comment the results

```{r eval=TRUE, echo=TRUE, include=TRUE}
resRR_PP_stIntI_cov = data.frame(resRR=RR_stIntI.BYM_cov, 
                       PP=unlist(PP_stIntI.BYM_cov),
                      SP_ID=RESP_DATAagg[,1])
# breakpoints
resRR_PP_stIntI_cov$resRRcat = cut(resRR_PP_stIntI_cov$resRR, breaks=c(min(resRR_PP_stIntI_cov$resRR), 
                  0.4, 0.6, 0.8, 1, 1.2, 1.4, 1.6, 
                  max(resRR_PP_stIntI_cov$resRR)),include.lowest = T)

resRR_PP_stIntI_cov$PPcat = cut(resRR_PP_stIntI_cov$PP, c(0, 0.2, 0.8, 1.00), include.lowest = TRUE)

map_RR_ST.IntI_cov = left_join(shape_file, resRR_PP_stIntI_cov, by = c("lad09cd" = "lad09cd"))
```


```{r mapRRstIntI, eval=TRUE, echo=TRUE, include=TRUE, fig.cap="Spatio-temporal model: Map of the residual RRs and posterior probabilities", , fig.width=12, fig.height=10}
p7 = ggplot() + geom_sf(data = map_RR_ST.IntI_cov) + aes(fill = resRRcat) +
  theme_bw() + scale_fill_brewer(palette = "PuOr") + 
  guides(fill=guide_legend(title="RR")) +  ggtitle("RR ST model Int I Covars") +
  theme(text = element_text(size=15), 
        axis.text.x = element_blank(), 
        axis.text.y = element_blank(), plot.title = element_text(size = 12, face = "bold")) 

p8 = ggplot() + geom_sf(data = map_RR_ST.IntI_cov) + aes(fill = PPcat) +
  theme_bw() +
  scale_fill_viridis(
    option = "plasma",
    name = "PP ST model Int I Covars",
    discrete = T,
    direction = -1,
    guide = guide_legend(
      title.position = 'top',
      reverse = T
    )) +  ggtitle("PP ST model Int I Covars") + theme(text = element_text(size=15), 
        axis.text.x = element_blank(), 
        axis.text.y = element_blank(), plot.title = element_text(size = 12, face = "bold"))

(p7 | p8)

# (p1|p2) / (p3|p4) / (p5|p6)
```


22. Plot the space-time interaction

```{r include=TRUE, fig.width=20, fig.height=10}
RESP_DATA_ST$intI_cov = stIntI.BYM.model_cov$summary.random$ID.space.time$mean

RESP_DATA_ST$intI_cov_cat = cut(RESP_DATA_ST$intI_cov,  breaks=c(-1,-0.05, 
                  -0.01, 0.01, 0.05, 1),include.lowest = T)
ggplot() +
  geom_sf(data = RESP_DATA_ST, aes(fill = intI_cov_cat))+ theme_bw() +  scale_fill_brewer(palette = "PuOr") + 
            guides(fill=guide_legend(title=NULL)) + 
            theme(text = element_text(size=20), 
                  axis.text.x = element_blank(), 
                  axis.text.y = element_blank()) +
facet_wrap(~ ID.time, ncol =8, labeller=labeller(ID.time=c("0"="2002","1"="2003","2"="2004","3"="2005","4"="2006","5"="2007","6"="2008","7"="2009","8"="2010","9"="2011", "10"="2012","11"="2013","12"="2014","13"="2015", "14"="2016","15"="2017"))) +
labs("")
```

23. Get a table of the hyperparameters. How do you interpret this table?

```{r eval=TRUE, echo=TRUE}

dat.hyper2 = 
  round(
  data.frame(median = stIntI.BYM.model_cov$summary.hyperpar[,4],
    LL = stIntI.BYM.model_cov$summary.hyperpar[,3], 
    UL = stIntI.BYM.model_cov$summary.hyperpar[,5]),
  digits = 3)

row.names(dat.hyper2) = 
  rownames(stIntI.BYM.model_cov$summary.hyperpar)

knitr::kable(dat.hyper2, caption = "Posterior median and 95% CrI of hyperparameters.") %>%  kable_styling(bootstrap_options = "striped", full_width = F, position = "center")
```

24. Compare the WAIC from the different models
```{r}

dat.WAIC = data.frame(model = c("Spatial", "SpatTemp no int", "SpatTemp noint covar", "SpatTemp typeI", "SpatTemp typeI covar"), 
                       WAIC = c(sBYM.model$waic$waic, stBYM.model$waic$waic,stBYM_cov$waic$waic, stIntI.BYM.model$waic$waic, stIntI.BYM.model_cov$waic$waic)
)

row.names(dat.WAIC) = NULL

knitr::kable(dat.WAIC, caption = "WAIC of the different models") %>%  kable_styling(bootstrap_options = "striped", full_width = F, position = "center")

```

ANSWER:
Although the spatial model has a lower WAIC it cannot be directly compared to the two spatio-temporal models, as it is essentially based on a different data set (i.e. data were aggregated removing the temporal component). The *fair comparison is between the two spatio-temporal models*, and we observe that the model with type I interaction performs better that the model without interaction.



